---
title: "EPI Analysis with Decision Tree"
output: 
  html_document:
    toc: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

# Load required libraries
suppressPackageStartupMessages({
  library(tidyverse)
  library(rpart)
  library(rpart.plot)
  library(DT)
})

# 1. Load and prepare the data
data <- read_csv("Data/epi2024_data.csv", show_col_types = FALSE) %>%
  select(country, BDH, ECS, FSH, APO, AGR, EPI) %>%
  drop_na()

# 2. Create EPI classes (Low/Medium/High)
data$EPI_Class <- cut(
  data$EPI,
  breaks = quantile(data$EPI, probs = c(0, 0.33, 0.66, 1), na.rm = TRUE),
  labels = c("Low", "Medium", "High"),
  include.lowest = TRUE
)

# 3. Train a simple decision tree
tree_model <- rpart(
  EPI_Class ~ BDH + ECS + FSH + APO + AGR,
  data = data,
  control = rpart.control(
    minsplit = 5,
    minbucket = 2,
    cp = 0.01,
    maxdepth = 3
  )
)

# 4. Get predictions
data$Predicted_Class <- predict(tree_model, type = "class")
accuracy <- mean(data$Predicted_Class == data$EPI_Class, na.rm = TRUE)
```

# EPI Decision Tree Analysis

This report presents a decision tree model for predicting Environmental Performance Index (EPI) classes based on environmental indicators.

## Decision Tree Visualization

```{r tree_plot, fig.width=12, fig.height=8}
rpart.plot(
  tree_model,
  type = 4,
  extra = 104,
  box.palette = list("#E41A1C", "#377EB8", "#4DAF4A"),
  fallen.leaves = TRUE,
  cex = 0.8,
  main = "EPI Decision Tree",
  tweak = 1.1,
  branch.lty = 3,
  shadow.col = "gray"
)
```

## Country Data

Search and sort the table below to explore the data:

```{r country_table}
# Prepare data for display
display_data <- data %>%
  select(
    Country = country,
    `EPI Score` = EPI,
    `EPI Class` = EPI_Class,
    `Predicted Class` = Predicted_Class,
    `Biodiversity (BDH)` = BDH,
    `Ecosystem Services (ECS)` = ECS,
    `Fish Stocks (FSH)` = FSH,
    `Air Pollution (APO)` = APO,
    `Agriculture (AGR)` = AGR
  ) %>%
  arrange(desc(`EPI Score`))

# Create interactive table
DT::datatable(
  display_data,
  rownames = FALSE,
  filter = 'top',
  options = list(
    pageLength = 10,
    autoWidth = TRUE,
    scrollX = TRUE
  )
) %>%
  formatRound(columns = c('EPI Score', 'Biodiversity (BDH)', 'Ecosystem Services (ECS)', 
                         'Fish Stocks (FSH)', 'Air Pollution (APO)', 'Agriculture (AGR)'), 
             digits = 1)
```

## Model Performance

The decision tree model achieves an accuracy of **`r round(accuracy * 100, 1)`%** in predicting EPI classes.

## How to Use This Tool

1. **Search**: Use the search box to find specific countries
2. **Sort**: Click on column headers to sort the data
3. **Filter**: Use the filter boxes below each column to filter the data
4. **Explore**: Compare actual vs predicted EPI classes to identify interesting patterns
